The Precambrian paleogeography of Laurentia

### 2021-03-10 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add Slave Province endnote and some minor edits

@@

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 billion years ago; Ga), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}). That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents. In contrast, South America and Africa are products of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca. 0.6 Ga \citep{Goscombe2019a}. Eurasia's constituent Proterozoic cratons have an even more recent history of amalgmation with the North China and South China cratons not arriving in their present relative position until ca. 0.15 Ga \citep{Van-der-Voo2015a,Torsvik2017a}. The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly. Rich geologic, paleomagnetic, and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}. Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen; THO  -- Trans-Hudson orogen; GFTZ -- Great Falls Tectonic Zone; MCR -- Midcontinent Rift). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a} while the diamond (not rated -- NR) poles are discussed in the text.}
\label{fig:Laurentia_map}
\end{figure*}

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. Throughout this chapter, I will utilize accretionary orogenesis in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia \citep{Staal2020a}. Collisional orogenesis will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure \citep{Staal2020a}. As discussed in \cite{Staal2020a}, there is ambiguity in this distinction as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale. Nevertheless, the categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.

@@

Overall, the collision of Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig. \ref{fig:Laurentia_map}). These collisional orogenies, including the Trans-Hudson orogeny, are interpreted to be associated with the initial assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}.

The Trans-Hudson orogeny is taken to be the terminal collision associated with closure of the Manikewan Ocean, a large oceanic tract that had previously separated the Superior and the North Atlantic provinces from the composite Slave+Rae+Hearne provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}; Fig. \ref{fig:Superior_Slave_recons}). The paleogeographic model of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia. In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography. In their model,  Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}). This core then subsequently grows to be a semi-supercontinent with India and Australia; however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.

@@

Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis. In the northwest, this accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margins of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the North Atlantic orogeny (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}). The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southeastern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

Following the Trans-Hudson orogeny, Laurentia's growth was dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts can be complicated by poor exposure in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Nevertheless, it is well-constrained that major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia which experienced metamorphic events between ca. 1.76 and 1.67 Ga (\citealp{Strickland2013a}; Fig. \ref{fig:Laurentia_map}). The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. However, it remains unclear whether the Mojave province should be considered a Archean province or a Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeastern Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

Laurentia's growth continued into the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. Continental arc magmatism is interpreted to have occurred associated with the Pinwarian orogeny in the northeast Grenville province in Labrador from ca. 1.52 to 1.46 Ga \citep{Gower2002a}. Accretionary orogenesis recorded in the Grenville province includes accretion of the Quebecia composite arc terrane to Laurentia ca. 1.43 to 1.37 Ga \citep{Groulier2020a}. Far to the southwest along the margin in northern New Mexico, metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths have been interpreted to indicate an interval of ca. 1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}. In the midcontinent region,  deformation and metamorphism of post-Mazatzal orogeny sedimentary rocks is constrained to have occurred ca. 1.49 to 1.46 Ga associated with the Baraboo orogeny \citep{Medaris2003a, Holm2019a}. Coveval Picuris-Baraboo-Pinwarian orogenesis is indicative of convergent tectonism along the entire length of the southeastern margin of Laurentia. It was in this active-margin, upper-plate setting that was of site of voluminous widespread plutonism from ca. 1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout the previously accreted Mesoproterozoic and Paleoproterozoic provinces. Known as the Granite-Rhyolite Province, these magmatic products extend from the southwestern United States to the Central Gneiss Belt of Ontario northeast of Georgian Bay \citep{Slagstad2009a}. This magmatism has been interpreted to be a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-lived active margin \citep{Bickford2015a}. Magmatic activity at the younger end of this range (ca. 1.37 Ga) is abundant in the Southern Granite-Rhyolite Province, suggesting a similar active margin setting \citep{Bickford2015a}. While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature and is consistent with evidence for accretionary orogenesis in the Picuris, Baraboo and Pinwarian orogens, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig2_Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
\label{fig:tectonic_history}
\end{figure*}

Accretionary orogenesis continued locally along the along the southeastern margin of Laurentia with the amalgamation and accretion of arcs and back-arcs associated with the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{Carr2000a,McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision and accretion of a previously rifted fragment of Laurentia that led to obduction of the Pyrites Complex ophiolite \citep{Carr2000a,McLelland2010a, Chiarenzelli2011a}. The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. An exception to this quiescence during the interval between the Shawinigan and Grenvillian orogenies is ca. 1.15 to 1.12 Ga orogenesis in the Llano uplift of the southern Laurentia margin \citep{Mosher1998a}. Llano orogenesis is interpreted to have resulted from collision of continental lithosphere along with an accreted arc \citep{Mosher1998a}. This orogenesis is earlier and temporally distinct from the Grenvillian orogeny, is only known from a limited spatial area, and is located in a region that experienced further orogenesis during the Grenvillian orogeny \citep{Grimes2004a}. Taken together, this context is suggestive of a microcontinent collision leading to Llano orogenesis prior to terminal Grenvillian continental collision. If this interpretation is correct, it would be similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (Fig. \ref{fig:tectonic_history}).

The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen throughout the orogen \citep{Carr2000a, Rivers2008a, Rivers2012a, Indares2020a}. Note that while the terms Grenvillian orogeny and Grenville belt have been used rather loosely in the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. Properly referring to the Grenvillian orogeny as distinct from the Elzevirian and Shawinigan orogenies enables the available constraints to be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}). Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a, Rivers2012a}. The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca. 1090 to 1030 Ma; \citealp{Rivers2008a}). Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca. 980 Ma \citep{Hynes2010a}.

In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior, leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}. This Midcontinent Rift is associated with the emplacement of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Cannon1994a, Swanson-Hysell2019a}.

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwestern Laurentia to the Ogilvie Mountains of northwestern Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g. \citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development. Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in southwestern Laurentia \citep{Yonkee2014a, Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting of the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia,  it suggests that significant lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia
break-up. One explanation is that Tonian extensional basin development was associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

>>>>

\usepackage{enotez}
\let\footnote=\endnote

@@

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 billion years ago; Ga), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}). That the vast majority of the present-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents. In contrast, South America and Africa are products of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca. 0.6 Ga \citep{Goscombe2019a}. Eurasia's constituent Proterozoic cratons have an even more recent history of amalgmation with the North China and South China cratons not arriving in their present relative position until ca. 0.15 Ga \citep{Van-der-Voo2015a,Torsvik2017a}. The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly. Rich geologic, paleomagnetic, and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}. Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen; THO  -- Trans-Hudson orogen; MCR -- Midcontinent Rift). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a} while the diamond (not rated -- NR) poles are discussed in the text.}
\label{fig:Laurentia_map}
\end{figure*}

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave\endnote{The term Slave Province is rather jarring when read for the first time given that it brings to mind evil human oppression. The geologic Slave Province as well as the eponymous Great Slave Lake get their name from a name given to the First Nation indigenous peoples of the Dene Group who are indigenous to the region. The origin of the name Slave is commonly explained as a French translation of the name given to the Dene people by the Cree people \citep{Britannica-Slave2017a}. Given its negative connotations, the name is not preferred now and there are efforts to remove it from Canadian place names \citep{Mandeville2016a}. Given that this name for the Archean geologic province is deeply entrenched in the geologic literature it will be used here.}+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. Throughout this chapter, I will utilize the term ``accretionary orogenesis'' in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia. ``Collisional orogenesis'' will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure. This usage follows \cite{Staal2020a} who discuss the ambiguity in such a division as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale. Nevertheless, this categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.

@@

These collisional orogenies, particularly the Trans-Hudson orogeny, are interpreted to be associated with assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}. The Trans-Hudson orogeny is taken to be the terminal collision associated with closure of the Manikewan Ocean, a large oceanic tract that had previously separated the Superior and the North Atlantic provinces from the composite Slave+Rae+Hearne provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}; Fig. \ref{fig:Superior_Slave_recons}). The paleogeographic model of \cite{Pehrsson2015a} posits the closure of the Manikewan Ocean not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia. In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography. In their model,  Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}). This core then subsequently grows to be a semi-supercontinent with India and Australia; however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.

Overall, the collision of Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the core of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).

@@

Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis. In the northwest, this accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margin of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the southern margin of the North Atlantic province (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}). The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

Following the Trans-Hudson orogeny, Laurentia's growth was dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts can be complicated by poor exposure in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Nevertheless, it is well-constrained that major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with arc-continent collision during the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia which experienced metamorphic events between ca. 1.76 and 1.67 Ga (\citealp{Strickland2013a}; Fig. \ref{fig:Laurentia_map}). The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. However, it remains unclear whether the Mojave province should be considered a Archean province or a Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended across the southeastern margin of Laurentia from the southwestern USA to eastern Canada where it is called the Labradorian orogeny. In eastern Canada, where the orogen was overprinted in the Grenvillian orogeny, it is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

Laurentia's growth continued into the Mesoproterozoic along the southeastern margin through further juvenile terrane and arc accretion. Continental arc magmatism is interpreted to have occurred associated with the Pinwarian orogeny in the northeast Grenville province in Labrador from ca. 1.52 to 1.46 Ga \citep{Gower2002a}. Accretionary orogenesis recorded in the Grenville province includes accretion of the Quebecia composite arc terrane to Laurentia ca. 1.43 to 1.37 Ga \citep{Groulier2020a}. Far to the southwest along the margin in northern New Mexico, metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths have been interpreted to indicate an interval of ca. 1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}. In the midcontinent region,  deformation and metamorphism of post-Mazatzal orogeny sedimentary rocks is constrained to have occurred ca. 1.49 to 1.46 Ga associated with the Baraboo orogeny \citep{Medaris2003a, Holm2019a}. Coveval Picuris-Baraboo-Pinwarian orogenesis is indicative of convergent tectonism along the entire length of the southeastern margin of Laurentia. This active-margin, upper-plate setting was the site of voluminous widespread plutonism from ca. 1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout the previously accreted Paleoproterozoic and Mesoproterozoic provinces. Known as the Granite-Rhyolite Province, these magmatic products extend from the southwestern United States to the Central Gneiss Belt of Ontario northeast of Georgian Bay (Fig. \ref{fig:Laurentia_map}; \citealp{Slagstad2009a}). This magmatism has been interpreted to be a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-lived active margin \citep{Bickford2015a}. Magmatic activity at the younger end of this range (ca. 1.37 Ga) is abundant in the Southern Granite-Rhyolite Province, suggesting a similar active margin setting \citep{Bickford2015a}. While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature and is consistent with evidence for accretionary orogenesis in the Picuris, Baraboo and Pinwarian orogens, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig2_Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past 1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) magmatism is also shown. The interpreted ages of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic (Table 2) and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
\label{fig:tectonic_history}
\end{figure*}

Accretionary orogenesis continued locally along the southeastern margin of Laurentia with the amalgamation and accretion of arcs and back-arcs associated with the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{Carr2000a,McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision and accretion of a previously rifted fragment of Laurentia that led to obduction of the Pyrites Complex ophiolite \citep{Carr2000a,McLelland2010a, Chiarenzelli2011a}. The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. An exception to this quiescence during the interval between the Shawinigan and Grenvillian orogenies is ca. 1.15 to 1.12 Ga orogenesis in the Llano uplift of the southern Laurentia margin \citep{Mosher1998a}. Llano orogenesis is interpreted to have resulted from collision of continental lithosphere along with an accreted arc \citep{Mosher1998a}. This orogenesis is earlier and temporally distinct from the Grenvillian orogeny, is only known from a limited spatial area, and is located in a region that experienced further orogenesis during the Grenvillian orogeny \citep{Grimes2004a}. Taken together, this context is suggestive of a microcontinent collision leading to Llano orogenesis prior to terminal Grenvillian continental collision. If this interpretation is correct, it would be similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (Fig. \ref{fig:tectonic_history}).

The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen \citep{Carr2000a, Rivers2008a, Rivers2012a, Indares2020a}. Note that while the terms Grenvillian orogeny and Grenville belt have been used rather loosely in the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. Properly referring to the Grenvillian orogeny as distinct from the Elzevirian and Shawinigan orogenies enables the available constraints to be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}). Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a, Rivers2012a}. The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca. 1090 to 1030 Ma; \citealp{Rivers2008a}). Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca. 980 Ma \citep{Hynes2010a}.

In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior, leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south (MCR in Fig. \ref{fig:Laurentia_map}; \citealp{Cannon1992b}). This Midcontinent Rift is associated with the emplacement of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Cannon1994a, Swanson-Hysell2019a}.

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwestern Laurentia to the Ogilvie Mountains of northwestern Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g. \citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development. Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in southwestern Laurentia \citep{Yonkee2014a, Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting off the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia,  it suggests that significant lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up. One explanation is that Tonian extensional basin development was associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence, there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
